\begin{Chapter}{Abstract}

This dissertation introduces the Modern AI Exposure Index (MAEI), a scenario-based structural index that measures how much each occupation in the United States overlaps with the capabilities of modern Generative Artificial Intelligence (GenAI). While earlier automation research focused on manual and routine labor, large language models like GPT-4 and Claude now target cognitive, analytical, and communicative tasks that were historically considered safe from technological displacement. To capture this shift, I built a five-stage computational pipeline anchored to two data foundations: \citet{Frey_Osborne_2017}'s historical automation probabilities for 702 occupations and the O*NET database (v30.1) containing over 130 occupational features across six domains. I engineered natural language features from O*NET task descriptions using BERT sentence embeddings (all-MiniLM-L6-v2) compressed via Principal Component Analysis to 20 orthogonal components. A Stacked Ensemble of XGBoost, Random Forest, and Support Vector Regression, with a Ridge meta-learner, reverse-engineered the 2013 structural relationship between occupational characteristics and automation risk, achieving a test $R^2$ of 0.6433 across 1,016 occupations. I then applied literature-calibrated capability multipliers to 68 features and a Broad Exposure Uplift formula to generate a counterfactual 2026 scenario. A 100-iteration Monte Carlo simulation bounded these assumptions with mathematical uncertainty. External validation against \citet{Felten_Raj_Seamans_2021}'s independently constructed AI Occupational Exposure dataset confirmed a strong positive Spearman correlation ($\rho = +0.764$, $p < 10^{-150}$) between the MAEI's pure exposure delta and the benchmark, while the historically anchored score showed the expected inverse correlation ($\rho = -0.357$), mathematically demonstrating the structural inversion of automation risk from manual to cognitive labor. A seven-scenario multiplier ablation study confirmed that the occupation rankings are structurally stable ($\rho$ range: 0.7636--0.7651) regardless of specific multiplier values. The MAEI provides a transparent, reproducible, and parameter-bounded map of occupational exposure to Generative AI.

\vspace{1em}
\noindent\textbf{Keywords:} \textit{artificial intelligence, occupational exposure, automation, machine learning, natural language processing, labor economics}

\end{Chapter}
